A flexible integer linear programming formulation for scheduling clinician on-call service in hospitals

### 2019-09-17 paper/paper.experiments.tex
Start fixing results (experiments) section

% JK: I think give an overview here of the experiments here
%     and what you're going to show in this section (and why)
\subsection{Implementation}
We developed a software package with a user interface that implements the above
linear program and allows configuration of clinicians at [ref \ref{???}], to be
% JK: what is this ref? Do you want to mention it is in Python?
used by the ID division at St.\ Michael's Hospital. The software was used to
generate the results in the following sections, using real data as well as
simulated data as input. All the following experiments were conducted on an
Intel Core i7-4770k CPU @ 3.50 GHz and 16 GB of RAM running 64-bit Windows 10.
The software was implemented using the COIN-OR Branch-and-Cut open source solver
% JK: "software" here is not clear to me, is it the solver for the ILP formulation?
version 2.9.9~\cite{johnjforrest_coin-or/cbc:_2019}.

@@

First, we evaluated the automatically-generated schedule by comparing it with the
% JK: I think it would be good to give a short name for manual and the proposed
%     data -- e.g. LP Solution as in Table 4 and Manual Generation.
%     Then check for consistency throughout.
%     While lack of synonyms can sound repetitive, it can help avoid confusion.
manually-created schedule. Specifically, we examined the adherence of each
schedule to the constraints presented in Table~\ref{tbl:constraint-summary}. As
shown in Table~\ref{tbl:constraints-comparison}, the optimal schedule generated
using the software satisfied all mandatory constraints. In contrast, the
% JK: "mandatory" = "hard"? if possible just use one.
manually-created schedule did not satisfy all mandatory constraints. In
particular, we see that the manual schedule assigned clinicians to multiple
consecutive blocks in all four years. Moreover, the manual schedule did not have
an equal distribution of weekends and holidays for all four years of data.
Considering all objectives, we see that LP outperforms the manually created
schedule in all four years, by accommodating almost all time-off requests and
ensuring that weekends are always assigned close to blocks.

@@

\subsection{Influence of Problem Complexity of Runtime}
Next, we examined the influence of the following four parameters on the
runtime of the ILP solver using simulated data:
number of clinicians;
number of services offered;
number of time-off requests per clinician per year;
time-horizon of the schedule.

% JK: Good to introduce the detault parameters first, then for each parameter explored,
%     Describe what varied.
We simulated a department with 10 clinicians offering two services, similar to the
department at St.\ Michael's Hospital. Each clinician was
configured with $[0, 5, 10, 15]$ non-overlapping block
% JK: \abs{BR} does not clearly imply number of blocks (could be confused with week indices),
%     so just leave out the LHS here I think.
% JK: What do you mean by non-overlapping? And why only increments of 5?
requests.
The effect of an increasing number of requests per clinician on the runtime of
the ILP solver is shown in Figure~\ref{fig:runtime-requests}.
The increase in the number of requests did not affect the runtime of
the ILP solver, indicating that it can accommodate a lot of flexibility in
clinician requests. Moreover, we see that all four runs were completed in under
1 second.
% JK: See somments below about trends and repeating simulations.

@@

% JK: Introduce what you did first and then the figure showing results (see above)
The effect of increasing the number of services %divisions %do you mean
%services?
on the runtime of the program is shown in Table~\ref{tbl:runtime-services-clinicians-comparison}.
We executed the algorithm for
$S = \{1, 2, 3\}$ total services and $C = \{10, 20, 30, 50\}$ clinicians in
total across all services. For 2 concurrent services, a roster of 30 or more
clinicians becomes impractical to schedule, as searching for a solution required
over 24 hours. We saw similar issues for a roster of 20 or more clinicians
assigned to a division with 3 concurrent services. However, when relaxing the
NCB constraint, we saw a great improvement in runtime for divisions with 2 and 3
services, and we were able to generate a schedule with upwards of 50 clinicians
in under 1.5 seconds.  %what does division mean? do you mean services?

>>>>

% JK: I think give an overview here of the experiments here
%     and what you're going to show in this section (and why)
In this section we compare the schedules created by solving the ILP formulation
given in Section~\ref{sec:methods} to schedules that were manually generated,
in terms of adherence to the hard and soft constraints outlined in Section~\ref{sec:problem}.
We also want to analyze the efficiency of the ILP approach in generating schedules
by its runtime on a variety of instances that may be found in the real-world.
Our goal is to show that the schedules provided by solving the ILP successfully
enforce all hard constraints and improve satisfiability of soft constraints, 
as well as to demonstrate that our ILP formulation can be used for a wide range
of configurations.

\subsection{Implementation}
We developed a Python software package with a user interface that implements the above
linear program and allows configuration of clinicians at~\cite{...}, to be
% JK: what is this ref? Do you want to mention it is in Python?
% TODO: I will add a ref to our Github repo here, once I clean it up and open source it
used by the ID division at St.\ Michael's Hospital. The software was used to
generate the results in the following sections, using real data as well as
simulated data as input. All the following experiments were conducted on an
Intel Core i7-4770k CPU @ 3.50 GHz with 16 GB of RAM running 64-bit Windows 10.
Our software package uses COIN-OR Branch-and-Cut open source solver
% JK: "software" here is not clear to me, is it the solver for the ILP formulation?
version 2.9.9~\cite{johnjforrest_coin-or/cbc:_2019}.

@@

First, we evaluated the ILP solution by comparing it with the
% JK: I think it would be good to give a short name for manual and the proposed
%     data -- e.g. LP Solution as in Table 4 and Manual Generation.
%     Then check for consistency throughout.
%     While lack of synonyms can sound repetitive, it can help avoid confusion.
manual generation, as in Table~\ref{tbl:2018-schedule-comparison}. Specifically, we examined the adherence of each
schedule to the constraints presented in Table~\ref{tbl:constraint-summary}. As
shown in Table~\ref{tbl:constraints-comparison}, the ILP solution satisfied all 
hard constraints. In contrast,
% JK: "mandatory" = "hard"? if possible just use one.
manual generation did not satisfy all hard constraints. In
particular, we see that the manual generation assigned clinicians to multiple
consecutive blocks in all four years. Moreover, the manual generation did not have
an equal distribution of weekends and holidays for all four years of data.
Considering all objectives, we see that the ILP solution outperforms manual generation
in all four years, by accommodating almost all time-off requests and
ensuring that weekends are always assigned close to blocks.

@@

\subsection{Influence of Problem Complexity on Runtime}
Next, we examined the influence of the following four parameters on the
runtime of the ILP solver using simulated data:
number of clinicians;
number of services offered;
number of time-off requests per clinician per year;
time-horizon of the schedule.

% JK: Introduce what you did first and then the figure showing results (see above)
The effect of increasing the number of clinicians and number of services %divisions %do you mean
%services?
on the runtime of the program is shown in Table~\ref{tbl:runtime-services-clinicians-comparison}.
We executed the algorithm for
$S = \{1, 2, 3\}$ total services and $C = \{10, 20, 30, 50\}$ clinicians in
total across all services. 
In a department providing a single service, increasing the number of clinicians
did not affect the runtime, and we were able to find an ILP solution in all four cases
within 1 second.
For 2 concurrent services, a roster of 30 or more
clinicians becomes impractical to schedule, as searching for a solution required
over 24 hours. We saw similar issues for a roster of 20 or more clinicians
assigned to a division with 3 concurrent services. However, when removing the
NCB constraint, we saw a great improvement in runtime for divisions with 2 and 3
services, and we were able to generate a schedule with upwards of 50 clinicians
in under 1.5 seconds.  %what does division mean? do you mean services?

% JK: Good to introduce the detault parameters first, then for each parameter explored,
%     Describe what varied.
For the remaining experiments, we simulated a department with 10 clinicians offering 
two services, similar to the department at St.\ Michael's Hospital. 
Each clinician was
configured with $\{0, 5, 10, 15\}$ non-overlapping block
% JK: \abs{BR} does not clearly imply number of blocks (could be confused with week indices),
%     so just leave out the LHS here I think.
% JK: What do you mean by non-overlapping? And why only increments of 5?
requests.
The effect of an increasing number of requests per clinician on the runtime of
the ILP solver is shown in Figure~\ref{fig:runtime-requests}.
The increase in the number of requests did not affect the runtime of
the ILP solver, indicating that it can accommodate a lot of flexibility in
clinician requests. Moreover, we see that all four runs were completed in under
1 second.
% JK: See somments below about trends and repeating simulations.

@@

### 2019-09-17 paper/tbl/2018_schedule_comparison.tex
Fix most chktex warnings
tables & figures cleanup
Start fixing results (experiments) section

@@

    \multicolumn{1}{c||}{\multirow{2}[1]{*}{Week \#}} & \multicolumn{3}{c||}{LP Solution} & \multicolumn{3}{c}{Historical Data} \\

@@

	\end{adjustbox}
	\caption{Comparison of schedules for 2018}

>>>>

  \caption{Comparison of automatically generated (ILP solution) and manually generated schedules for 2018. In the ILP solution there are rigid 2-week block assignments, unlike manual generation that often assigns 3-4 weeks in a row to a single clinician. Moreover, in the ILP solution we see for each block either the HIV or ID clinician was assigned weekend coverage, indicating improved Block-Weekend Adjacency.}%
  % JK: could you clarify in the caption what we should take away from this table?

@@

    \multicolumn{1}{c||}{\multirow{2}[1]{*}{Week \#}} & \multicolumn{3}{c||}{ILP Solution} & \multicolumn{3}{c}{Manual Generation} \\

@@

	\end{adjustbox}\\[1em]
	\footnotesize\raggedright
	The ``HIV/ID'' columns represent the assignments for the two concurrent services offered at the department. The ``Weekend'' column represents the assignments for weekend coverage in both services. Different colours and letters are used to distinguish different clinicians.
  % JK: should use a footnote to define the meaning of
  %     - HIV / ID
  %     - Weekend
  %     - Colors & Letters
  %     so that the reader can understand the table basically without reading the paper.
